---
title: "Modeling"
author: "Madalyn Young"
date: "2025-03-03"
output: 
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes # makes the toc move along
    code_folding: "show"  # Use "hide" to collapse code by default
editor_options: 
  chunk_output_type: inline
---

```{r}
pacman::p_load(tidyverse, scales, dplyr, corrr, janitor, tidyr, psych, readr, lubridate, rpart, rpart.plot, caret, C50, sf, maps, dbscan, geosphere)
```


# Load Original Data
```{r}
#original data
CustomerProfileData <- read.csv("Data/customer_profile.csv")  
TransactionalData <- read.csv("Data/transactional_data.csv")
AddressZipData <- read.csv("Data/customer_address_and_zip_mapping.csv")
DeliveryCostData <- read.csv("Data/delivery_cost_data.csv")
```



```{r}
#clean the address data
#This is code from Imogen

# Split the column
AddressZipData <- AddressZipData |>
  separate(full.address, into = c("ZIP", "City", "State Name", "State Short", 
                                  "County","Code", "Latitude", "Longitude"), sep = ",")

AddressZipData$Latitude <- as.numeric(AddressZipData$Latitude)

AddressZipData$Longitude <- as.numeric(AddressZipData$Longitude)
```

   
```{r}
#Pivot wide the cost data
#aggregated  transaction data to join to customer table

# aggregate transaction data by customer_number and year
#sum the ordered cases and gallons by customer number and year
#this table is set up so each customer number has  two rows, one for 2023, one for 2024. Each column is sum of ordered cases/loaded cases. delivered cases in that year 
aggregated_cost <- TransactionalData |>
  group_by(CUSTOMER_NUMBER, YEAR) |>
  summarize(orderedCases = sum(ORDERED_CASES),
            orderedGallons = sum(ORDERED_GALLONS))



#The code pivots the database above to have one row per customer and a column for each cases/gallons ordered for each year
aggregated_cost_wide <- aggregated_cost |>
  pivot_wider(
    names_from = YEAR, 
    values_from = c(orderedCases, 
                    orderedGallons),
    names_sep = "_"
  )
```

```{r}
#clean Customer Profile Data
  CustomerProfileData <-  CustomerProfileData %>% 
  mutate(
    Entity_ID = case_when(
      is.na(PRIMARY_GROUP_NUMBER) ~ CUSTOMER_NUMBER,  # If PRIMARY_GROUP_NUMBER is NA, use CUSTOMER_NUMBER
      TRUE ~ PRIMARY_GROUP_NUMBER),
    ON_BOARDING_DATE = mdy(ON_BOARDING_DATE),
    FIRST_DELIVERY_DATE = mdy(FIRST_DELIVERY_DATE),
    ON_BOARDING_YEAR = year(ON_BOARDING_DATE),
    FIRST_DELIVERY_YEAR = year(FIRST_DELIVERY_DATE))

char_col <- sapply(CustomerProfileData, is.character)
CustomerProfileData[char_col] <- lapply(CustomerProfileData[char_col], as.factor)

logical_cols <- sapply(CustomerProfileData, is.logical)
CustomerProfileData[logical_cols] <- lapply(CustomerProfileData[logical_cols], as.numeric)

#CustomerProfileData$ZIP_CODE <- #as.character(CustomerProfileData$ZIP_CODE)

#remove the customer where their on_boarding date was first delivery date was before the onboarding date (1 customer)
CustomerProfileData <- CustomerProfileData %>% 
  filter(FIRST_DELIVERY_DATE>=ON_BOARDING_DATE)
```



# Location Cleaning/EDA


```{r}
# join customer data to wide cost data and Address data
Main_Customer_Data1 <- CustomerProfileData %>% 
  left_join(aggregated_cost_wide, by = "CUSTOMER_NUMBER") %>% 
  left_join(AddressZipData, by = c("ZIP_CODE"="zip")) 


Main_Customer_Data1 <- Main_Customer_Data1 %>% 
  mutate(excludeinclude = case_when(
    year(FIRST_DELIVERY_DATE) == 2023 & (orderedCases_2023 == 0 & orderedGallons_2023 == 0 )~"exclude",
    year(FIRST_DELIVERY_DATE) == 2024 & (orderedCases_2024 == 0 & orderedGallons_2024 == 0 )~"exclude",
    TRUE ~"include")) 
  
```

```{r}
#cluster the addresses and calculate the centroid for each cluster
##Multiple centroids
set.seed(123)

kmeans_result <- kmeans(Main_Customer_Data1[,c("Longitude", "Latitude")], centers = 4)

Main_Customer_Data1$cluster <- as.factor(kmeans_result$cluster)


centroids <- Main_Customer_Data1 %>% 
  group_by(cluster) %>% 
  summarize(centroid_lon = mean(Longitude), centroid_lat = mean(Latitude))

```

```{r}

haversine_distance <- function(lon1, lat1, lon2, lat2) {
  distHaversine(c(lon1, lat1), c(lon2,lat2))/1609.34 
}# converts meters to miles

#Join main customer data to the clusters created above
Main_Customer_Data1 <- Main_Customer_Data1 %>% 
  left_join(centroids, by = "cluster")


Main_Customer_Data1 <- Main_Customer_Data1 %>% 
  mutate(
    distance_to_centroid = mapply(haversine_distance, Main_Customer_Data1$Longitude, Main_Customer_Data1$Latitude, Main_Customer_Data1$centroid_lon, Main_Customer_Data1$centroid_lat)
  )

```






```{r}
us_map <- map_data("usa")
states <- map_data("state")

ggplot()+
  geom_polygon(data = states, aes(x = long, y = lat, group = group), fill = "gray90", color = "black", size = 0.3) +
  geom_point(data = Main_Customer_Data1, aes(x = Longitude, y = Latitude), color = Main_Customer_Data1$cluster, size = 3)+
  geom_point(data = centroids, aes(x=centroid_lon, y=centroid_lat), color = "red", size = 5, shape = 4) + 
   coord_cartesian(xlim = c(-105, -65))+ theme_minimal()
```


-- make a map plot with only outlet customers to see their spread



# Create New Main Data Source
```{r}
Main_Customer_Data2 <- Main_Customer_Data1 %>% 
  mutate(across(c(orderedCases_2023, orderedCases_2024, orderedGallons_2023, orderedGallons_2024), ~ replace_na(.x, 0)),
         total_ordered = (orderedCases_2024 + orderedGallons_2024 +orderedCases_2023 + orderedGallons_2023),
         total_ordered_2023 = (orderedCases_2023 + orderedGallons_2023),
         total_ordered_2024 = (orderedCases_2024 + orderedGallons_2024)) %>% 
  group_by(Entity_ID) %>% 
  summarize(FREQUENT_ORDER_TYPE = FREQUENT_ORDER_TYPE[which.max(tabulate(match(FREQUENT_ORDER_TYPE, unique(FREQUENT_ORDER_TYPE))))],
            COLD_DRINK_CHANNEL = first(COLD_DRINK_CHANNEL),
            TRADE_CHANNEL = first(TRADE_CHANNEL),
            SUB_TRADE_CHANNEL = first(SUB_TRADE_CHANNEL),
            FIRST_DELIVERY_DATE = min(FIRST_DELIVERY_DATE),
            FIRST_DELIVERY_YEAR = min(FIRST_DELIVERY_YEAR),
            ON_BOARDING_DATE = min(ON_BOARDING_DATE),
            ON_BOARDING_YEAR = min(ON_BOARDING_YEAR),
            LOCAL_MARKET_PARTNER = LOCAL_MARKET_PARTNER[which.max(tabulate(match(LOCAL_MARKET_PARTNER,unique(LOCAL_MARKET_PARTNER))))],
            CO2_CUSTOMER =  CO2_CUSTOMER[which.max(tabulate(match(CO2_CUSTOMER, unique(CO2_CUSTOMER))))],
            hasOutlet = first(case_when(is.na(PRIMARY_GROUP_NUMBER)~ 0,TRUE ~1)),
            numberOfOutlets = sum(case_when(is.na(PRIMARY_GROUP_NUMBER)~ 1,TRUE ~1)),
            wellPerformingOutlet = sum(case_when((orderedGallons_2023 + orderedCases_2023) >= 400 ~ 1, (orderedGallons_2024 + orderedCases_2024) >=400 ~ 1, TRUE ~ 0)),
            
            hasOrderedCases = as.integer(mean(case_when((orderedCases_2023 + orderedCases_2024)>0 ~1, TRUE ~ 0))>0,1,TRUE~0),
            
            proprtionOfCases = sum(orderedCases_2023, orderedCases_2024)/ sum(total_ordered),
            
            GeoSpread = n_distinct(ZIP),
            most_common_zip = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(ZIP),
              ZIP[which.max(tabulate(match(ZIP, unique(ZIP))))]), 
            largest_zip = if_else(
              numberOfOutlets == 1,
              first(ZIP),
              ZIP[which.max(total_ordered)]
            ),
            
            most_common_city = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(City),
              City[which.max(tabulate(match(City, unique(City))))]), 
            largest_city = if_else(
              numberOfOutlets == 1,
              first(City),
              City[which.max(total_ordered)]
            ),
            
            most_common_state = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(`State Name`),
              `State Name`[which.max(tabulate(match(`State Name`, unique(`State Name`))))]), 
            largest_state = if_else(
              numberOfOutlets == 1,
              first(`State Name`),
              `State Name`[which.max(total_ordered)]
            ),
            
            most_common_region = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(cluster),
              cluster[which.max(tabulate(match(cluster, unique(cluster))))]), 
            largest_region = if_else(
              numberOfOutlets == 1,
              first(cluster),
              cluster[which.max(total_ordered)]
            ),
            
            most_common_distance = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(distance_to_centroid),
              distance_to_centroid[which.max(tabulate(match(distance_to_centroid, unique(distance_to_centroid))))]), 
            largest_distance = if_else(
              numberOfOutlets == 1,
              first(distance_to_centroid),
              distance_to_centroid[which.max(total_ordered)]
            ),
            
            total_ordered_2023 = sum(total_ordered_2023),
            total_ordered_2024 = sum(total_ordered_2024),
            
            ) %>% 
     mutate(ThreshBins = case_when(
    total_ordered_2023 < 1000 | total_ordered_2024 <  1000 ~ "<1K",
    (total_ordered_2023 >= 1000 & total_ordered_2023 < 10000) | (total_ordered_2024 >= 1000 & total_ordered_2024 < 10000) ~ "1K-10K",
     (total_ordered_2023 >= 10000 & total_ordered_2023 < 100000) | (total_ordered_2024 >= 10000 & total_ordered_2024 < 100000) ~ "10K-100K",

    total_ordered_2023 > 100000  | total_ordered_2024 > 100000 ~ ">100K"),
    percentChangeYOY = ((total_ordered_2024) - (total_ordered_2023))/(total_ordered_2023)) %>% 
    mutate(excludeinclude = case_when(
    FIRST_DELIVERY_YEAR == 2023 & (total_ordered_2023 == 0 )~"exclude",
    FIRST_DELIVERY_YEAR == 2024 & (total_ordered_2024 == 0 )~"exclude",
    TRUE ~"include")) %>% 
  filter(excludeinclude == 'include')

            
```


 Change Main Datasour
 - remove 50 customers with onborading date and no orders in 2023 and 2024 - DONE
 - exclude customer with onboarding date after first delivery date - DONE
 - add a column on boarding year - DONE
 - bring in case indicator
   - has case - DONE
   - proportion of cases to total orders - DONE 
   - remove the TF and both ordered columns - DONE
- remove delivery loaded columns - DONE
- remove well performing outlet (keep the 2 year one -> the one that says 400 gallons in 2023 OR 2024) - DONE
- remove two year threshmet - DONE
- zip code 
   - log and lat column - DONE
   - geospread - maybe for outlet customers, how many regions/clusters they are in - DONE
   - most_common_Zip - DONE
   - calcuate the difference between the density area and long lat of the customer (need to EDA to find the density area ) - DONE
   
   
# Break out Main_Data_Source by parent and child customers

```{r}
Main_Customer_Data_HAS_OUTLET <- Main_Customer_Data1 %>% 
  mutate(across(c(orderedCases_2023, orderedCases_2024, orderedGallons_2023, orderedGallons_2024), ~ replace_na(.x, 0)),
         total_ordered = (orderedCases_2024 + orderedGallons_2024 +orderedCases_2023 + orderedGallons_2023),
         total_ordered_2023 = (orderedCases_2023 + orderedGallons_2023),
         total_ordered_2024 = (orderedCases_2024 + orderedGallons_2024)) %>% 
  group_by(Entity_ID) %>% 
  summarize(FREQUENT_ORDER_TYPE = FREQUENT_ORDER_TYPE[which.max(tabulate(match(FREQUENT_ORDER_TYPE, unique(FREQUENT_ORDER_TYPE))))],
            COLD_DRINK_CHANNEL = first(COLD_DRINK_CHANNEL),
            TRADE_CHANNEL = first(TRADE_CHANNEL),
            SUB_TRADE_CHANNEL = first(SUB_TRADE_CHANNEL),
            FIRST_DELIVERY_DATE = min(FIRST_DELIVERY_DATE),
            FIRST_DELIVERY_YEAR = min(FIRST_DELIVERY_YEAR),
            ON_BOARDING_DATE = min(ON_BOARDING_DATE),
            ON_BOARDING_YEAR = min(ON_BOARDING_YEAR),
            LOCAL_MARKET_PARTNER = LOCAL_MARKET_PARTNER[which.max(tabulate(match(LOCAL_MARKET_PARTNER,unique(LOCAL_MARKET_PARTNER))))],
            CO2_CUSTOMER =  CO2_CUSTOMER[which.max(tabulate(match(CO2_CUSTOMER, unique(CO2_CUSTOMER))))],
            hasOutlet = first(case_when(is.na(PRIMARY_GROUP_NUMBER)~ 0,TRUE ~1)),
            numberOfOutlets = sum(case_when(is.na(PRIMARY_GROUP_NUMBER)~ 1,TRUE ~1)),
            wellPerformingOutlet = sum(case_when((orderedGallons_2023 + orderedCases_2023) >= 400 ~ 1, (orderedGallons_2024 + orderedCases_2024) >=400 ~ 1, TRUE ~ 0)),
            
            hasOrderedCases = as.integer(mean(case_when((orderedCases_2023 + orderedCases_2024)>0 ~1, TRUE ~ 0))>0,1,TRUE~0),
            
            proprtionOfCases = sum(orderedCases_2023, orderedCases_2024)/ sum(total_ordered),
            
            GeoSpread = n_distinct(ZIP),
            most_common_zip = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(ZIP),
              ZIP[which.max(tabulate(match(ZIP, unique(ZIP))))]), 
            largest_zip = if_else(
              numberOfOutlets == 1,
              first(ZIP),
              ZIP[which.max(total_ordered)]
            ),
            
            most_common_city = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(City),
              City[which.max(tabulate(match(City, unique(City))))]), 
            largest_city = if_else(
              numberOfOutlets == 1,
              first(City),
              City[which.max(total_ordered)]
            ),
            
            most_common_state = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(`State Name`),
              `State Name`[which.max(tabulate(match(`State Name`, unique(`State Name`))))]), 
            largest_state = if_else(
              numberOfOutlets == 1,
              first(`State Name`),
              `State Name`[which.max(total_ordered)]
            ),
            
            most_common_region = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(cluster),
              cluster[which.max(tabulate(match(cluster, unique(cluster))))]), 
            largest_region = if_else(
              numberOfOutlets == 1,
              first(cluster),
              cluster[which.max(total_ordered)]
            ),
            
            most_common_distance = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(distance_to_centroid),
              distance_to_centroid[which.max(tabulate(match(distance_to_centroid, unique(distance_to_centroid))))]), 
            largest_distance = if_else(
              numberOfOutlets == 1,
              first(distance_to_centroid),
              distance_to_centroid[which.max(total_ordered)]
            ),
            
            total_ordered_2023 = sum(total_ordered_2023),
            total_ordered_2024 = sum(total_ordered_2024),
            
            ) %>% 
     mutate(ThreshBins = case_when(
    total_ordered_2023 < 1000 | total_ordered_2024 <  1000 ~ "<1K",
    (total_ordered_2023 >= 1000 & total_ordered_2023 < 10000) | (total_ordered_2024 >= 1000 & total_ordered_2024 < 10000) ~ "1K-10K",
     (total_ordered_2023 >= 10000 & total_ordered_2023 < 100000) | (total_ordered_2024 >= 10000 & total_ordered_2024 < 100000) ~ "10K-100K",

    total_ordered_2023 > 100000  | total_ordered_2024 > 100000 ~ ">100K"),
     percentChangeYOY = ((total_ordered_2024) - (total_ordered_2023))/(total_ordered_2023)) %>% 
    mutate(excludeinclude = case_when(
    FIRST_DELIVERY_YEAR == 2023 & (total_ordered_2023 == 0 )~"exclude",
    FIRST_DELIVERY_YEAR == 2024 & (total_ordered_2024 == 0 )~"exclude",
    TRUE ~"include")) %>% 
  filter(excludeinclude == 'include') %>% 
  filter(numberOfOutlets >1) 
```


```{r}

outletGraph <- Main_Customer_Data1 %>% 
  group_by(Entity_ID) %>% 
  mutate(numberOfOutlets = sum(case_when(is.na(PRIMARY_GROUP_NUMBER)~ 1,TRUE ~1))) %>% 
  filter(numberOfOutlets==5)


ggplot()+
  geom_polygon(data = states, aes(x = long, y = lat, group = group), fill = "gray90", color = "black", size = 0.3) +
  geom_point(data =outletGraph, aes(x = Longitude, y = Latitude), color = (outletGraph$Entity_ID), size = 3)+
  facet_wrap(~Entity_ID, scales = "free") +
  theme_minimal()
```

```{r}
Main_Customer_Data_NO_OUTLET <-  Main_Customer_Data1 %>% 
  mutate(across(c(orderedCases_2023, orderedCases_2024, orderedGallons_2023, orderedGallons_2024), ~ replace_na(.x, 0)),
         total_ordered = (orderedCases_2024 + orderedGallons_2024 +orderedCases_2023 + orderedGallons_2023),
         total_ordered_2023 = (orderedCases_2023 + orderedGallons_2023),
         total_ordered_2024 = (orderedCases_2024 + orderedGallons_2024)) %>% 
  group_by(Entity_ID) %>% 
  summarize(FREQUENT_ORDER_TYPE = FREQUENT_ORDER_TYPE[which.max(tabulate(match(FREQUENT_ORDER_TYPE, unique(FREQUENT_ORDER_TYPE))))],
            COLD_DRINK_CHANNEL = first(COLD_DRINK_CHANNEL),
            TRADE_CHANNEL = first(TRADE_CHANNEL),
            SUB_TRADE_CHANNEL = first(SUB_TRADE_CHANNEL),
            FIRST_DELIVERY_DATE = min(FIRST_DELIVERY_DATE),
            FIRST_DELIVERY_YEAR = min(FIRST_DELIVERY_YEAR),
            ON_BOARDING_DATE = min(ON_BOARDING_DATE),
            ON_BOARDING_YEAR = min(ON_BOARDING_YEAR),
            LOCAL_MARKET_PARTNER = LOCAL_MARKET_PARTNER[which.max(tabulate(match(LOCAL_MARKET_PARTNER,unique(LOCAL_MARKET_PARTNER))))],
            CO2_CUSTOMER =  CO2_CUSTOMER[which.max(tabulate(match(CO2_CUSTOMER, unique(CO2_CUSTOMER))))],
            hasOutlet = first(case_when(is.na(PRIMARY_GROUP_NUMBER)~ 0,TRUE ~1)),
            numberOfOutlets = sum(case_when(is.na(PRIMARY_GROUP_NUMBER)~ 1,TRUE ~1)),
            wellPerformingOutlet = sum(case_when((orderedGallons_2023 + orderedCases_2023) >= 400 ~ 1, (orderedGallons_2024 + orderedCases_2024) >=400 ~ 1, TRUE ~ 0)),
            
            hasOrderedCases = as.integer(mean(case_when((orderedCases_2023 + orderedCases_2024)>0 ~1, TRUE ~ 0))>0,1,TRUE~0),
            
            proprtionOfCases = sum(orderedCases_2023, orderedCases_2024)/ sum(total_ordered),
            
            GeoSpread = n_distinct(ZIP),
            most_common_zip = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(ZIP),
              ZIP[which.max(tabulate(match(ZIP, unique(ZIP))))]), 
            largest_zip = if_else(
              numberOfOutlets == 1,
              first(ZIP),
              ZIP[which.max(total_ordered)]
            ),
            
            most_common_city = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(City),
              City[which.max(tabulate(match(City, unique(City))))]), 
            largest_city = if_else(
              numberOfOutlets == 1,
              first(City),
              City[which.max(total_ordered)]
            ),
            
            most_common_state = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(`State Name`),
              `State Name`[which.max(tabulate(match(`State Name`, unique(`State Name`))))]), 
            largest_state = if_else(
              numberOfOutlets == 1,
              first(`State Name`),
              `State Name`[which.max(total_ordered)]
            ),
            
            most_common_region = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(cluster),
              cluster[which.max(tabulate(match(cluster, unique(cluster))))]), 
            largest_region = if_else(
              numberOfOutlets == 1,
              first(cluster),
              cluster[which.max(total_ordered)]
            ),
            
            most_common_distance = if_else(
              numberOfOutlets == 1,  # If only one location, take that ZIP
              first(distance_to_centroid),
              distance_to_centroid[which.max(tabulate(match(distance_to_centroid, unique(distance_to_centroid))))]), 
            largest_distance = if_else(
              numberOfOutlets == 1,
              first(distance_to_centroid),
              distance_to_centroid[which.max(total_ordered)]
            ),
            
            total_ordered_2023 = sum(total_ordered_2023),
            total_ordered_2024 = sum(total_ordered_2024),
            
            ) %>% 
     mutate(ThreshBins = case_when(
    total_ordered_2023 < 501 | total_ordered_2024 <  500 ~ "<500",
    (total_ordered_2023 >= 501 & total_ordered_2023 < 1001) | (total_ordered_2024 >= 501 & total_ordered_2024 < 1001) ~ "500-1K",
     (total_ordered_2023 >= 1001 & total_ordered_2023 < 2001) | (total_ordered_2024 >= 1001 & total_ordered_2024 < 2001) ~ "1K-2K",

    total_ordered_2023 >= 2001  | total_ordered_2024 > 2001 ~ ">2K"),
    percentChangeYOY = ((total_ordered_2024) - (total_ordered_2023))/(total_ordered_2023)) %>% 
    mutate(excludeinclude = case_when(
    FIRST_DELIVERY_YEAR == 2023 & (total_ordered_2023 == 0 )~"exclude",
    FIRST_DELIVERY_YEAR == 2024 & (total_ordered_2024 == 0 )~"exclude",
    TRUE ~"include")) %>% 
  filter(excludeinclude == 'include') %>% 
  filter(numberOfOutlets ==1) 
```

```{r}
NoOutletTrain <- Main_Customer_Data_NO_OUTLET %>% 
  filter((total_ordered_2023>=400 | total_ordered_2024>=400))
```


```{r}
C50Model_NoOutlet <- C5.0(as.factor(ThreshBins)~ FREQUENT_ORDER_TYPE + COLD_DRINK_CHANNEL + SUB_TRADE_CHANNEL + LOCAL_MARKET_PARTNER + most_common_zip + CO2_CUSTOMER + as.factor(ON_BOARDING_YEAR) + most_common_distance + most_common_region + hasOrderedCases + proprtionOfCases  ,NoOutletTrain, CF = .25, earylStopping = FALSE, noGlobalPruning = FALSE)
```

```{r}
plot(C50Model_NoOutlet)

```

